---
title: "Combining Dependent p-values"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Supplementary R code for reproducing the examples in the paper ``Combining Dependent p-values: Methods and Properties from a Practical Perspective'' by Gasparin M., Goeman J.J., Heller R. and Solari A.


## Set up

Install and load the required R packages:

```{r, results='hide', message=F, warning=F, error=F, comment=NA}
require("readxl") || install.packages("readxl")
require("pmerge") || install.packages("pmerge")
require("Iso") || install.packages("Iso")
```

## Domain-Level Analysis of the Functional Observational Battery

The functional observational battery (FOB) is a biological screening assay consisting of $m=25$ endpoints grouped into six domains, each reflecting different aspects of neurological function, and is used to assess neurophysiological effects in animals following toxin exposure.

```{r}
# Load data
options(timeout = 300)

download.file(
  "https://pasteur.epa.gov/uploads/10.23719/d-gsv0/Moser_1995_FOB.zip",
  "moser.zip",
  mode = "wb"
)

unzip("moser.zip")

df <- as.data.frame(read_excel("E220RANK_Acute.xlsx"))

# Restrict to dose groups 0 and 1500 at time point 4
dat_raw <- df[
  df$DOSE_CODE %in% c(0, 1500) &
    df$TIME_OF_TESTING == 4,
  -c(1:2, 4, 33:38)
]

# Convert variables to numeric
dat_num <- data.frame(
  lapply(dat_raw, function(x) as.numeric(as.character(x))),
  check.names = FALSE
)

# Group endpoints by functional domain
endpoints_by_domain <- list(
  Autonomic = c(
    "LACRIMATION", "SALIVATION", "PALPEBRAL_CLOSURE...6",
    "PUPIL_RESPONSE", "URINATION", "DEFECATION"
  ),
  Activity = c(
    "REARS", "HOME_CAGE_POSTURE"
  ),
  Excitability = c(
    "EASE_OF_REMOVAL", "EASE_OF_HANDLING", "AROUSAL",
    "CLONIC_CONVULSIONS", "TONIC_CONVULSIONS"
  ),
  Neuromuscular = c(
    "GAIT_SCORE", "RIGHTING_REFLEX", "FORELIMB_GRIP",
    "HINDLIMB_GRIP", "FOOT_SPLAY"
  ),
  Sensorimotor = c(
    "TAIL_PINCH_RESPONSE", "CLICK_RESPONSE",
    "TOUCH_RESPONSE", "APPROACH_RESPONSE"
  ),
  Physiological_measures = c(
    "BODY_WEIGHT", "BODY_TEMPERATURE", "PILOERECTION"
  ),
  Other = c(
    "DOSE_CODE", "MOBILITY_SCORE", "VOCALIZATIONS",
    "PALPEBRAL_CLOSURE...10"
  )
)

# Select domains for analysis
domains_keep <- c(
  "Autonomic",
  "Activity",
  "Excitability",
  "Neuromuscular",
  "Sensorimotor",
  "Physiological_measures"
)

cols_keep <- c("DOSE_CODE", unlist(endpoints_by_domain[domains_keep]))
dat <- dat_num[, intersect(cols_keep, names(dat_num)), drop = FALSE]
```

Score-free test statistic

```{r}
Tmax <- function(Y1, Y2, K = 4) {
  # Ordered categories assumed to be 1,2,3,4 (K=4 by default).
  # Works even if some categories are not observed.
  
  # drop NAs
  Y1 <- Y1[!is.na(Y1)]
  Y2 <- Y2[!is.na(Y2)]
  
  # basic checks
  if (length(Y1) == 0 || length(Y2) == 0) return(NA_real_)
  if (any(!Y1 %in% 1:K) || any(!Y2 %in% 1:K)) {
    stop("Y1 and Y2 must take values in {1,2,...,K}.")
  }
  
  # tabulate counts for all categories 1,...,K
  r1 <- tabulate(Y1, nbins = K)
  r2 <- tabulate(Y2, nbins = K)
  t  <- r1 + r2
  
  # keep only pooled-observed categories
  idx <- t > 0
  c   <- sum(idx)
  
  n1 <- sum(r1)
  n2 <- sum(r2)
  if (n1 == 0 || n2 == 0) return(NA_real_)
  
  # group indicator
  W <- rep(c(0, 1), times = c(n1, n2))
  
  # if only one category is observed overall
  if (c <= 1) return(0)
  
  # stochastic ordering check
  lhs <- cumsum(r2[idx]) / n2
  rhs <- cumsum(r1[idx]) / n1
  
  if (sum(lhs >= rhs) == c) {
    
    M <- matrix(0, nrow = c - 1, ncol = c)
    E <- matrix(0, nrow = c - 1, ncol = n1 + n2)
    F <- numeric(c - 1)
    
    for (i in 1:(c - 1)) {
      M[i, ] <- rep(c(0, 1), times = c(i, c - i))
      E[i, ] <- rep(c(M[i, ], M[i, ]),
                    times = c(r1[idx], r2[idx]))
      F[i]   <- suppressWarnings(cor(E[i, ], W, method = "pearson"))
    }
    
    T <- max(F, na.rm = TRUE)
    
  } else {
    
    # isotonic regression branch
    if (!exists("pava", mode = "function")) {
      stop("pava() not found. Please source pava.R or load a package providing pava().")
    }
    
    y   <- r2[idx] / t[idx]
    w   <- t[idx]
    iso <- pava(y, w)
    
    Yt  <- rep(c(iso, iso),
               times = c(r1[idx], r2[idx]))
    T   <- suppressWarnings(cor(Yt, W, method = "pearson"))
  }
  
  as.numeric(T)
}
```



For each endpoint, marginal counts of severity scores
(1–4) for adverse effects observed are reported by exposure group (0 and 1.5g/kg). Endpoints are grouped into functional domains.

```{r}
n <- sum(dat$DOSE_CODE == 0)
N <- nrow(dat)
m <- ncol(dat) - 1
Z <- as.matrix(dat[, -1])
perms <- combn(1:N, n)
B <- ncol(perms)

# ---- Marginal counts (scores 1–4)

freq_fun <- function(x) {
  t(apply(x, 2, function(col)
    tabulate(factor(col, levels = 1:4), nbins = 4)))
}

g1 <- Z[1:8, ]
g2 <- Z[9:16, ]

f1 <- freq_fun(g1)
f2 <- freq_fun(g2)

mfreq <- cbind(f1, f2)
colnames(mfreq) <- c(1:4, 1:4)
rownames(mfreq) <- colnames(Z)

Tstat <- matrix(NA, nrow = B, ncol = m)

for (b in 1:B) {
  Zperm <- rbind(Z[perms[, b], ], Z[-perms[, b], ])
  Tstat[b, ] <- sapply(
    1:ncol(Zperm),
    function(i) Tmax(Zperm[1:n, i], Zperm[(n + 1):N, i])
  )
}

pval <- apply(Tstat, 2, function(t) mean(t >= t[1]))

data.frame(mfreq, pval)
```

Domain-wise combined $p$-values.

```{r}
# ---- Domain-wise combination 
domains <- endpoints_by_domain[-length(endpoints_by_domain)]
domains$All <- 1:25
domains <- c(tail(domains, 1), head(domains, -1))

comb_funs <- c("Bonferroni","Simes","Hommel+","Fisher","Average","Median","Cauchy","Harmonic")

res <- matrix(NA, nrow = length(domains), ncol = length(comb_funs) * 2)
colnames(res) <- c(comb_funs, paste0(comb_funs, "U"))
rownames(res) <- paste(names(domains), " (",sapply(domains,length),")", sep="")

for (d in seq_along(domains)) {
  
  Zsel <- Z[, domains[[d]]]
  Tstat <- matrix(NA, nrow = B, ncol = ncol(Zsel))
  
  for (b in 1:B) {
    Zperm <- rbind(Zsel[perms[, b], ], Zsel[-perms[, b], ])
    Tstat[b, ] <- sapply(
      1:ncol(Zperm),
      function(i) Tmax(Zperm[1:n, i], Zperm[(n + 1):N, i])
    )
  }
  
  pval <- apply(Tstat, 2, function(t) mean(t >= t[1]))
  
  Pmat <- apply(Tstat, 2, function(t) {
    sapply(seq_along(t), function(i) mean(t >= t[i]))
  })
  
  # ---- Joint permutation p-values ----
  
  minP <- apply(Pmat, 1, min)
  res[d,1] <- mean(-minP >= -minP[1])
  
  SimesP <- apply(Pmat, 1, function(x) min(sort(x) * length(x)/(1:length(x))))
  res[d,2] <- mean(-SimesP >= -SimesP[1])
  
  HommelP <- apply(Pmat, 1, function(x) min(sort(x) * length(x)/(1:length(x))))
  res[d,3] <- mean(-HommelP >= -HommelP[1])
  
  FisherP <- apply(Pmat, 1, function(x) -sum(log(x)))
  res[d,4] <- mean(FisherP >= FisherP[1])
  
  AverageP <- rowMeans(Pmat)
  res[d,5] <- mean(-AverageP >= -AverageP[1])
  
  MedianP <- apply(Pmat, 1, function(x) sort(x)[ceiling(length(x)/2)])
  res[d,6] <- mean(-MedianP >= -MedianP[1])
  
  CauchyP <- apply(Pmat, 1, function(x) atan(mean(tan(pi*(x-1/2))) )/pi+1/2 )
  res[d,7] <- mean(-CauchyP >= -CauchyP[1])
  
  HarmonicP <- apply(Pmat, 1, function(x) mean(x^(-1))^(1/(-1)) )
  res[d,8] <- mean(-HarmonicP >= -HarmonicP[1])
  
  # ---- Marginal combination ----
  
  res[d,9]  <- min(1, min(pval) * length(pval))
  res[d,10] <- min(1, min(sort(pval) * length(pval)/(1:length(pval))))
  res[d,11] <- pSimes(pval, method = "G")
  res[d,12] <- pmean(pval, r = 0, dependence = "A")
  res[d,13] <- min(1, 2 * mean(pval))
  res[d,14] <- min(1, 2 * sort(pval)[ceiling(length(pval)/2)])
  res[d,15] <- NA
  res[d,16] <- ifelse(length(pval)>2, pharmonic(pval, method = "H2", censor = TRUE), NA )
}

res
```

